Value-Suppressing Uncertainty Maps

### 2017-03-31 paper/template.tex
Fixed typo

Using a dataset from a group of virologists interested in the population dynamics of viral mutation~\cite{o2012conditional}, we visualized the variability in viral genomes across 25 populations of the simian immunodeficiency virus (SIV). Next generation sequencing (NGS) allows biologists to collect and analyze large amounts of genomics data. However, these techniques often create data quality issues, as they must align large numbers of potentially ambiguous ``reads'' of relatively small numbers of base pairs. In the dataset here, it is important not only to identify hotspots in the viral genome (locations with high rates of mutation or variability), but to not be distracted by false positives. The traditional bivariate map (\figref{fig:viral2d}) affords the analysis of only a few, large hot spots. aExtending the ``confidence fog'' metaphor used in viral genome browsers like LayerCake~\cite{correll2015layercake}, VSUMs (\figref{fig:viralVsum}) enable biologists to explore interesting patterns while not being distracted by noise. Smaller scale ``hot spots'' that all be mapped to the same magenta color in the traditional map are visible as bright orange lines. Pink hot spots towards the end of the genome (past reference nucleotide $9,000$) that could be plausibly considered potential regions of interest in the 2D map, are encoded as highly uncertain and of ambiguous value in the VSUMs. In fact, these regions have systematically low coverage; the lack of reliable data for these regions should discourage analysts from giving their apparently pattern equal consideration from the  stronger signals earlier in the genome.

>>>>

Using a dataset from a group of virologists interested in the population dynamics of viral mutation~\cite{o2012conditional}, we visualized the variability in viral genomes across 25 populations of the simian immunodeficiency virus (SIV). Next generation sequencing (NGS) allows biologists to collect and analyze large amounts of genomics data. However, these techniques often create data quality issues, as they must align large numbers of potentially ambiguous ``reads'' of relatively small numbers of base pairs. In the dataset here, it is important not only to identify hotspots in the viral genome (locations with high rates of mutation or variability), but to not be distracted by false positives. The traditional bivariate map (\figref{fig:viral2d}) affords the analysis of only a few, large hot spots. Extending the ``confidence fog'' metaphor used in viral genome browsers like LayerCake~\cite{correll2015layercake}, VSUMs (\figref{fig:viralVsum}) enable biologists to explore interesting patterns while not being distracted by noise. Smaller scale ``hot spots'' that all be mapped to the same magenta color in the traditional map are visible as bright orange lines. Pink hot spots towards the end of the genome (past reference nucleotide $9,000$) that could be plausibly considered potential regions of interest in the 2D map, are encoded as highly uncertain and of ambiguous value in the VSUMs. In fact, these regions have systematically low coverage; the lack of reliable data for these regions should discourage analysts from giving their apparently pattern equal consideration from the  stronger signals earlier in the genome.
